Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-01 AppendixI.Rmd
intro

**Figure A4**: Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities **(a)** Richness, **(b)** Shannon and **(c)** Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3 \@ref(fig:DivTaxo).

**Figure A5**: Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes. \@ref(fig:DivFun).

**Figure A6**: Summarizing plots of the functional PCA analysis defining communities functional space. **(a)** Plot of the samples from the BRIDGE functional database in the two dimensional main plan. **(b)** bow plot of the PCA eignevalues in percentage. **(c)** Plot of the 7 functional traits sed in the PCA analysis (Leaf and stem traits, and maximum species height). \@ref(fig:PCA_graphs).

**Figure A7**: Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3. \@ref(fig:RedFun).

@@

# **Figure A4**: Richness, Shannon and Simpson taxonomic diversity trajectories

```{r DivTaxo, out.width = '100%', fig.asp = 1.5/3, echo=FALSE, fig.cap="Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities (ref:Bolda) Richness, (ref:Boldb) Shannon and (ref:Boldc) Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")

TaxoTraj(CompleteTaxo)

# **Figure A5**: Rao diversity trajectories with all plots display

@@

# **Figure A6**: Simpson and Rao Linera regrassion summary

@@

# **Figure A7**: Functional PCA

@@

# **Figure A8**: Overall functional redundancy

>>>>

**Figure A4**: Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes. \@ref(fig:DivFun).

**Figure A5**: Summarizing plots of the functional PCA analysis defining communities functional space. **(a)** Plot of the samples from the BRIDGE functional database in the two dimensional main plan. **(b)** bow plot of the PCA eignevalues in percentage. **(c)** Plot of the 7 functional traits sed in the PCA analysis (Leaf and stem traits, and maximum species height). \@ref(fig:PCA_graphs).

**Figure A6**: Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3. \@ref(fig:RedFun).

@@

# **Figure A4**: Rao diversity trajectories with all plots display

@@

# **Figure A5**: Simpson and Rao Linera regrassion summary

@@

# **Figure A6**: Functional PCA

@@

# **Figure A7**: Overall functional redundancy

### 2018-07-01 WholePlotTrajectories.Rmd
intro

    To disentangle the ecological processes driving communities response to disturbance we analysed the taxonomic and functional diversity trajectories following a logging and thinning disturbance gradient.
    Specifically we examined over 30 years the trajectories of communities taxonomic richness and evennes and functional composition, diversity and redundancy based on 7 leaf, stem and life history traits.
    Trajectories highlighted communities taxonomic resilience, maintaining their initial differences in composition, and functional resilience through fast functional trajectory common to all communities in the functional space.
    The IDH consistently predicted communities functional trajectories according to the disturbance intensity, but poorly represented their taxonomic trajectories, blurried by the alteration of the functional redundancy and the recruitment of infrequent, shade tolerant species slowed by competitive exclusion processes.

    Although communities functioning had recovered after 30 years their taxonomic diversity and composition in infrequent species and their functional redundancy remained altered.
    these results acknowledged the need of decades-long recovery cycles to ensure communities complete resilience, and questioned communities maintenance after repeated disturbance.

@@

In tropical forest, ecological communities are constantly re-shaped by natural disturbance events that change both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions like competition among species [@Chesson2000;@Herault2018]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to disturbance [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through forest structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
These structural parameters were thereafter sucessfully modeled, giving important insights into the recovery of ecosystems processes and services [@Herault2018]. 
However the response of forests diversity is still unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014, @Liang2016] and would be most probably impacted by the changes induced by disturbance [@Baraloto2012a].

In the short-term, moderate disturbance may lead to positive impacts on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
Still, validations of the IDH remain scarce in the long term and mainly rely on the analysis of species richness that gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More complete analysis would besides encompass communities composition that is crucial for conservation issues, and abundance distribution to reveal the ecological rules structuring communities [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accounting for species biological attributes and role in the ecosystem directly links ecosystems biodiversity, functioning and environmental constraints [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
The functional trait-based approach focused on major traits related to species ecology and performance was sucessfully adopted [@Diaz2005;@Villeger2008a], for example to reveal in tropical rainforests the deterministic processes that foster after disturbance the fast growing species with efficient resources acquisition [@Molino2001;@Haddad2008;@Ruger2009].
This basic deterministic process entails a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
This shift is translated into consistent trajectories of key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction life history traits (seed mass and maximum height) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Besides communities functional composition and diversity, a last key aspect of communities response to disturbance is the changes in functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016].
High functional redundancy, as in the very diverse tropical forests [@Bellwood2006], mitigates the impacts of species removal on ecosystem functioning and determines communities resilience [@Trenbath1999; @Elmqvist2003; @Diaz2005].
High functional redundancy also increases the functional overlap among species and would foster neutral stochastic processes [@Gravel2006].

To grasp all facets of communities response to disturbance we examined here the taxonomic and functional trajectories in diversity and composition [@Lohbeck2015;@Guariguata2001].
These trajectories would highlight the recovery of communities initial composition, diversity and functioning and the underlying ecological processes.
They would clarify the tenants in the long term of the Intermediate Disturbance Hypothesis debated in tropical forests, and provide indications for future adaptive conservation strategies [@Adler2007].
We monitored over 30 years the response of 75 ha of neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed functional traits and species maximum height to draw the trajectories over time of communities taxonomic and functional composition and diversity. Specifically, we (i) questioned the coupling between taxonomic and functional response to disturbance and identified the underlying assembly processes, (ii) clarified the validity of the IDH in the long term for tropical forest and (iii) questioned the completeness of communities recovery regarding communities functional redundancy.

@@

## Inventories protocol and dataset collection {#protocols}

@@

To counter taxonomic uncertainties due to the variability of botanical identification levels (in space) and protocols (in time), the taxonomic composition and diversity analysis were conducted at the genus level, *i.e.* referring to the genus of observed or trialed botanical names.
Trajectories of communities taxonomic and functional variations in composition after disturbance were followed in a two-dimensional NMDS ordination space of the 30 years monitored.
Two NMDS were conducted to map either taxonomic flora inventories or communities functional composition based on the 7 leaf, stem and life history traits (without seed mass classes).
In both cases the NMDS were performed using abundance-based (Bray-Curtis) dissimilarity measures. 
Trajectories along time were reported through the distance between successive inventories and the reference inventories in 1989, *i.e.* 2-3 years after disturbance, when the uncertainty degree did not exceed 30% of undetermined trees. 
Taxonomic trajectories were measured through the euclidean distance between inventories in the multivariate space of all inventories species.
Functional trajectories were measured through the sum of euclidean distances among traits density distributions between inventories.
To compensate the intrinsinc difference among plots, the trajectories corresponded to the euclidean differences along time with the reference inventory in 1989.
Univariate trajectories of the leaf, stem and life-history traits were also visualized with the community weighted means (CWM) [@Diaz2007;@Garnier2004]. 
Species seed mass corresponded to 5 classes of increasing mass, seed mass trajectories were therefore reported as the proportion of each class in the inventories.

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973].
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), recomended for diversity studies [@Patil1982;@Tothmeresz1995;@Marcon2015]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and life traits.

The impacts of initial disturbance were first tested with the spearman rank correlation between the extremum of taxonomic and functional metrics reached over the 30 years and the initial %AGB removed. 
Then they were analysed through the linear correlations between Simpson and Rao diversities and the initial %AGB removed at 10, 20 and 30 years after disturbance.

@@

## Communities Diversity

From 1989 (2-3 years after disturbance) to 2015 (28-29 years after disturbance), 828388 individual trees and 591 botanical species pertaining to 223 genus and 64 botanical families were recorded. 
For undisturbed plots Richness and taxonomic evenness (Shannon and Simpson diversities) remained stable over the 30 years monitored.
In disturbed communities the taxonomic richness increased after low disturbance intensity, reaching a maximum gain of 14 botanical genera (plot 3 from treatment 2) while it followed unimodal trajectories after intense disturbance, decreasing for ten years before recovering to pre-disturbance values.
In all disturbed plots the taxonomic evenness (Shannon and Simpson diversities) increased, following unimodal trajectories with a maximum, reached after around 20 years, positively correlated to the disturbance treatment ($\rho_{spearman}^{Shannon}=0.86$, and $\rho_{spearman}^{Simpson}=0.89$).
Return towards initial evenness values was beginning after 30 years except for two T3 plots (plots 8 and 12) which evenness still increased, suggesting similar but delayed trajectories \@ref(fig:IDHplot).

Trajectories of communities functional diversity were examined through the Rao diversity based on the 7 leaf, stem and life history traits (to the exception of seed mass).
The plot 7 from treatment 1 displayed a constantly outlying diversity and was removed from the graphical representation for better readability (see appendix for full graphs).
In undisturbed plots the functional diversity remained stable along the 30 years while in disturbed plots it followed unimodal trajectories with a return towards initial values that strated around 20 years after disturbance.

@@

```{r IDHplot, out.width = '100%', echo=FALSE, fig.asp = 1/1.5,fig.cap="Upper panels, Trajectories of the Simpson taxonomic diversity (ref:Bolda) and Rao functional diversity (ref:Boldb) over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \\%AGB removed and Simpson (ref:Boldc) and Rao (ref:Boldd) diversities 10, 20 and 30 years after disturbance.", fig.env="figure*"}
load("DB/SimpsonDraw");load("DB/FunctionalTraj_ForGraphs")
load("DB/SimpsonIDH");load("DB/RaoIDH");load("DB/LostAGB")

par(mfrow=c(2,2),mar=c(3,3,2,1),oma=c(2,1,2,3),no.readonly = T)
plotDiv(SimpsonTaxo)
mtext("Simpson diversity",side=3,adj=0,line=1)
plotDiv(CompleteFun,remove=TRUE)
mtext("Rao diversity",side=3,adj=0,line=1)
mtext("Years since disturbance",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.28,0),xpd=NA,legend=c("T0","T1","T2","T3"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment")

@@

>> Si on met que richesse et evenness, pour functional c'est de l'evenness (Rao = order2)

@@

The taxonomic differences among communities, marked before disturbance by the distinct starting points of the NMDS, were maintained throughout communities recovery trajectories.
More than commonly thought post-disturbance trajectories depended on communities initial composition, that partly determined the pool of rectuitable species and that constrained the converge towards the recovery of the initial composition [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a]. 
The resilience of communities taxonomy suggested the existence of multiple stable equilibria, here the pre-disturbance communities, as assumed for highly diverse and productive ecosystems [@Chase2003].

@@

The humped-shaped trajectories of communities along time demonstrated the enhanced functional evennes and changes in functional composition following disturbance before a return towards initial, lower values.
In accordance with the IDH, disturbance entailed significant temporal variability with the emergence of ecological processes different from those rueling undisturbed forest.
The increase in functional evenness went with significant functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998; @Wright2004; @Reich2014].
Because the pre-disturbance survivor trees mirror the initial communities [@Herault2018], the functional changes relied upon recruited trees and the enhanced growth and survival of previously infrequent species and functional types.
Pioneers outcompeted other species in the post-disturbance environment of non limiting resources and alleviated competition.
Before disturbance and over communities recovery time though, these species would be excluded by long-lived, more resistant and shade tolerant species.

@@

The recovery of the most infrequent species followed the lottery recruitment rules [@Busing2002] and was hampered by the increasing competition among species whih entailed long term recovery cycles [@Trenbath1999;@Elmqvist2003;@Diaz2005].

The rapid decrease of the functional redundancy within the initial functional space, correlated with the disturbance intensity, illustrated the space and resources made available after disturbance.
Following the lottery recruitment, communities taxonomic richness increased firsthand along the enhanced recruitment rate but decreased thereafter, following the emergence of exclusive competition. 
After 30 years the recovery of the initial redundancy structure was underway but remained unachieved.
This alteration of the functional redundancy meant a lower resilience of pre-disturbance communities and higher chances to see the persistence of disturbance-specific communities, with less species and more pioneers [@Haddad2008; @Burslem2000; @Martin2013].
Besides the hampered recovery of rare species increased the risks to loose cornerstone specie, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].
Apart from the functional characteristics considered here, infrequent species mighthave unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance entailing the resilience of their functioning and taxonomic composition with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for  20 years after disturbance through the enrichment of the communities with pioneers and light-demanding species,in accordance with the IDH.
The IDH, though, poorly predicted the disturbance impact on communities taxonomic richness and evenness that were blurried by the emergence of competitive exclusion along time.
The resilience of tropical forests proved consistent although spread over several decades. 
Still, the disturbance impact on communities redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Gourlet-Fleury2005].
As the trajectories highlighted the recruitment processes proved central for communities response to disturbance and closer focus on demographical drivers of communities response would clarify the fate of the future forests.

>>>>

    To disentangle the ecological processes driving communities response to disturbance we analysed the taxonomic and functional diversity trajectories following a logging and thinning disturbance gradient.
    Specifically we examined over 30 years the trajectories of communities taxonomic richness and evennes and functional composition, diversity and redundancy based on 7 leaf, stem and life history traits.
    Trajectories highlighted the cyclic recovery of communities taxonomic composition, maintaining the pre-disturbance differences, as well as functional composition and diversity.
    The IDH predicted communities functional response to disturbance that resulted from competition processes determined by the disturbance intensity.
    The IDH however poorly represented the taxonomic response that relied upon lottery recruitment and recruitment limitation which hampered the recovery of infrequent, shade tolerant species.

    After disturbance communities functioning followed a cyclic recovery and the initial composition differences among plots were maintained.
    Although consistent the recovery of communities functioning and taxonomic composition and diversity remained unachieved after 30 years which acknowledged the need of decades-long recovery cycles to ensure a complete recovery, and questioned communities sustainability after repeated disturbance.

@@

In tropical forests, ecological communities are constantly re-shaped by natural disturbance events that change both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions like competition among species [@Chesson2000;@Herault2018]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to disturbance [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through forest structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
These structural parameters were thereafter sucessfully modeled, giving important insights into the recovery of ecosystems processes and services [@Herault2018]. 
However the response of forests diversity is still unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014, @Liang2016] and would be most probably impacted by the changes induced by disturbance [@Baraloto2012a].
In the short-term, moderate disturbance may lead to positive impacts on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity when the disturbance regime is moderate but significant [@Molino2001;@Kariuki2006a;@Berry2008a]. 
Along time the IDH would translate into a variety of post-disturbance trajectories shaped by the different ecological processes at stake.

Validations of the IDH though remain scarce in the long term and mainly rely on the analysis of species richness.
Species richness though gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016]: more complete analysis would encompass communities composition that is crucial for conservation issues, and abundance distribution that reveals the ecological processes involved [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accounting for species biological attributes and assessing their role in the ecosystem would directly link communities diversity to the ecosystem functioning and to its environmental constraints [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
In that respect, the functional trait-based approach that focused on major traits related to species ecology and performance was sucessfully adopted [@Diaz2005;@Villeger2008a].
For example the functoinal approach revealed in tropical rainforests the deterministic processes fostering fast growing species with efficient resources acquisition after disturbance [@Molino2001;@Haddad2008;@Ruger2009].
Disturbance therefore entails a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
This shift is translated into consistent trajectories of key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction life history traits (seed mass and maximum height) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually a complete overview of communities response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
High functional redundancy like in the very diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities [@Trenbath1999; @Elmqvist2003; @Diaz2005].
High functional redundancy also increases the functional overlap among species and would foster neutral stochastic processes [@Gravel2006].

To grasp all facets of communities response to disturbance we examined here the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy [@Lohbeck2015;@Guariguata2001].
These trajectories would highlight the recovery of communities initial characteristics and the underlying ecological processes, and provide indications for future adaptive conservation strategies [@Adler2007].
They would clarify the tenants of the Intermediate Disturbance Hypothesis in the long term, debated in tropical forests, and explicit the variety of post-disturbance trajectories defined by the disturbance intensity.
We monitored over 30 years the response of 75 ha of neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed functional traits and species maximum height to draw the trajectories over time of communities taxonomic and functional composition, richness, evenness and redundancy. Specifically, we (i) questioned the recovery of communities taxonomic and functional characteristics and identify the underlying assembly processes, (ii) clarified the validity of the IDH in the long term for tropical forest and elucidated its declination into different trajectories, and (iii) questioned the length of complete communities recovery.

@@

## Inventories protocol and dataset collection

@@

To counter taxonomic uncertainties due to the variability of botanical identification levels (in space) and protocols (in time), the taxonomic composition and diversity analysis were conducted at the genus level, *i.e.* referring to the genus of observed or trialed botanical names.
Trajectories of communities taxonomic and functional variations in composition after disturbance were followed in a two-dimensional NMDS ordination space of the 30 years monitored.
Two NMDS were conducted to map either taxonomic flora inventories or communities functional composition based on the 7 leaf, stem and life history traits (without seed mass classes).
In both cases the NMDS were performed using abundance-based (Bray-Curtis) dissimilarity measures. 
Trajectories along time were reported through the distance between successive inventories and the reference inventories in 1989, *i.e.* 2-3 years after disturbance, when the uncertainty degree did not exceed 30% of undetermined trees. 
for the taxonomix composition trajectories were measured by the euclidean distance between inventories in the multivariate space of all inventories species.
For the functoinal composition trajectories were measured by the sum for all traits of the euclidean distances between the traits density distributions of successive inventories.
To compensate the intrinsinc difference among plots, the trajectories corresponded to the euclidean differences along time with the reference inventory in 1989.
Univariate trajectories of the leaf, stem and life-history traits were also visualized with the community weighted means (CWM) [@Diaz2007;@Garnier2004]. 
Species seed mass corresponded to 5 classes of increasing mass, seed mass trajectories were therefore reported as the proportion of each class in the inventories.

The taxonomic diversity was reported through species richness and the evenness, *i.e* the Hill number translation of the Simpson index [@Hill1973].
These two indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recomended for diversity studies [@Patil1982;@Tothmeresz1995;@Marcon2015]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and life traits.

The impacts of initial disturbance were tested with the spearman rank correlation between the extremum of taxonomic and functional metrics reached over the 30 years and the initial %AGB removed. 
The were besides analysed through the linear correlations between Simpson and Rao diversities and the initial %AGB removed at 10, 20 and 30 years after disturbance.

@@

## Communities richness and evenness

From 1989 (2-3 years after disturbance) to 2015 (28-29 years after disturbance), 828388 individual trees and 591 botanical species pertaining to 223 genus and 64 botanical families were recorded. 
For undisturbed plots taxonomic Richness and Evenness (Simpson diversity) remained stable over the 30 years monitored.
In disturbed communities after low disturbance intensity the taxonomic richness increased, reaching a maximum gain of 14 botanical genera (plot 3 from treatment 2).
After intense disturbance the taxonomic richness followed unimodal trajectories, decreasing for ten years after disturbance before recovering to pre-disturbance values.
In all disturbed plots the taxonomic evenness (Simpson diversity) followed unimodal trajectories with first an increase until a maximum reached after around 20 years that was positively correlated to the disturbance treatment ($\rho_{spearman}^{Shannon}=0.86$, and $\rho_{spearman}^{Simpson}=0.89$).
The return towards initial values of communities evenness was beginning after 30 years except for two T3 plots (plots 8 and 12) which evenness still increased, suggesting similar but delayed trajectories \@ref(fig:DivTaxo).

```{r DivTaxo, out.width = '100%', fig.asp = 1.5/3, echo=FALSE, fig.cap="Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities (ref:Bolda) Richness, (ref:Boldb) Shannon and (ref:Boldc) Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")

TaxoTraj(CompleteTaxo)

The plot 7 from treatment 1 displayed a constantly outlying functional evenness and was removed from the graphical representation for better readability (see appendix for full graphs).
In undisturbed plots the functional evenness remained stable along the 30 years while in disturbed plots it followed unimodal trajectories with a return towards initial values starting around 20 years after disturbance.

```{r FunDiv, out.width = '100%', echo=FALSE, fig.asp = 1/1.5,fig.cap="Trajectories of the Rao functional diversity over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals.", fig.env="figure"}
load("DB/FunctionalTraj_ForGraphs")

plotDiv(CompleteFun,remove=TRUE)
mtext("Rao diversity",side=3,adj=0,line=1)
mtext("Years since disturbance",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.28,0),xpd=NA,legend=c("T0","T1","T2","T3"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment")
```

@@

```{r IDHplot, out.width = '100%', echo=FALSE, fig.asp = 1/1.5,fig.cap="Upper panels, Trajectories of the taxonomic evenness (Simpson diversity) (ref:Bolda) and Rao functional diversity (ref:Boldb) over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \\%AGB removed and Simpson (ref:Boldc) and Rao (ref:Boldd) diversities 10, 20 and 30 years after disturbance.", fig.env="figure*"}
load("DB/SimpsonDraw");load("DB/FunctionalTraj_ForGraphs")
load("DB/SimpsonIDH");load("DB/RaoIDH");load("DB/LostAGB")

par(mfrow=c(1,2),mar=c(3,3,2,1),oma=c(2,1,2,3),no.readonly = T)
plotDiv(SimpsonTaxo)
mtext("Simpson diversity",side=3,adj=0,line=1)
plotDiv(CompleteFun,remove=TRUE)
mtext("Rao diversity",side=3,adj=0,line=1)
mtext("Years since disturbance",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.28,0),xpd=NA,legend=c("T0","T1","T2","T3"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment")

@@

@@

The taxonomic differences among communities, marked before disturbance by the distinct starting points of the NMDS, were maintained throughout communities recovery trajectories.
More than commonly thought post-disturbance trajectories depended on communities initial composition, that partly determined the pool of recruitable species and that constrained the converge towards the recovery of the initial composition [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a]. 
The resilience of communities taxonomy suggested the existence of multiple stable equilibria, here the pre-disturbance communities, as assumed for highly diverse and productive ecosystems [@Chase2003].

@@

The humped-shaped trajectories of communities along time illustrated the processes enhancing the functional evennes and the changes in functional composition following disturbance before communities return towards initial composition and lower evenness.
Disturbance entailed significant variability in the communities through the emergence of ecological processes different from thoses involvemnt before disturbance, in accordance with the IDH.
The increase in functional evenness went with significant functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998; @Wright2004; @Reich2014].
Because the pre-disturbance survivor trees mirror the initial communities [@Herault2018], the functional changes relied upon recruited trees and the enhanced growth and survival of previously infrequent species and functional types.
Disturbance vacated environmental niches of high light, space and nutrient availability, filled by competitive pioneers that became dominant and determined the functional characteristics of the community [@Grime1998].
Before disturbance and over communities recovery time though, these species would be excluded by long-lived, more resistant and shade tolerant species.

@@

The recovery of the most infrequent species followed the lottery recruitment rules [@Busing2002] and was hampered by the increasing competition among species, implying long term recovery [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Following disturbance the space and resources made available corresponded to a rapid decrease of the functional redundancy within the initial functional space.
The high light, space and nutrient niches were rapidly filled by species, mainly the most dominant and frequent ones.
Thereafter competitive exclusion emerged following the filling of the niches and limited the recruitment of infrequent species [@Busing2002].
The recovery of communities taxonomic richness and evenness relied upon the recovery of the initial functional redudancy.
Although underway 30 years after disturbance the recovery remained unachieved and after 30 years the recovery it was underway but remained unachieved.
This alteration of the functional redundancy meant a lower resilience of pre-disturbance communities and higher chances to see the persistence of disturbance-specific communities, with less species and more pioneers [@Haddad2008; @Burslem2000; @Martin2013].
Besides the slowed recovery of rare species increased the risks to loose cornerstone specie, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].
Apart from the functional characteristics considered here, infrequent species mighthave unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance allowing the resilience of their functioning and taxonomic composition with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for  20 years after disturbance through the enrichment of the communities with pioneers and light-demanding species,in accordance with the IDH.
The IDH, though, poorly predicted the disturbance impact on communities taxonomic richness and evenness that were blurried by the emergence of competitive exclusion along time.
The resilience of tropical forests proved consistent although spread over several decades. 
Still, the disturbance impact on communities redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Gourlet-Fleury2005].
As the trajectories highlighted the recruitment processes proved central for communities response to disturbance and closer focus on demographical drivers of communities response would clarify the fate of the future forests.
